---
title: "Survival analysis"
author: "Guillaume Nony & Saïd Hamdi"
date: "2023-10-25"
output: pdf_document
---

## Step1 - Installation packages and loading dataset

R packages used :

-   'survival' =\> compute survival analyses
-   'survminer' =\> summarize and visualize the results of survival analysis.

```{r}
# install.packages(c("survival", "survminer"))
library("survival")
library("survminer")
```

We are going to work on the lung cancer dataset. The lung cancer dataset is available in the survival package. The dataset is composed of the following data :

-   inst: Institution code
-   time: Survival time in days
-   status: censoring status 1=censored, 2=dead
-   age: Age in years
-   sex: Male=1 Female=2
-   ph.ecog: ECOG performance score (from 0=active to 5=dead)
-   ph.karno: Karnofsky performance score (from 0=dead to 100=normal)
-   pat.karno: Karnofsky performance score as rated by patient
-   meal.cal: Calories consumed at meals
-   wt.loss: Weight loss in last six months

For more detail on ph. ecog and ph. karno : <https://ecog-acrin.org/resources/ecog-performance-status/>

the data is clean and consistent. We don't need to make any corrections.

```{r}
head(lung) # look at the dataset
```

The data is clean and consistent. We don't need to make any corrections.

## Step 2 - Compute survival curves per sex with survfit()

The function survfit() is included in survival package. The Kaplan-Meier estimator is a non-parametric statistic used to estimate the survival function from lifetime data.

```{r}
fit <- survfit(Surv(time, status) ~ sex, data = lung)  
print(fit)
```

### Access to the variables in the function survfit()

```{r}
str(fit)
```

The main arguments are :

-   n: total number of observations in each curve (M/F).
-   time: the time points on the curve.
-   n.risk: the number of observations at risk at time t.
-   n.event: the number of events that occurred at time t.
-   n.censor: the number of censored observations at time t.
-   surv : estimate of survival at time t.
-   std.err : std error of the cumulative hazard
-   strata: counts nb of time points belong to each curve. Each curve is also given a name.
-   lower,upper: lower and upper confidence limit.

Remark on censored observations : An event of interest (death, relapse) may not be observed during the study time. It produces a censored observations (patient lost, ...).

```{r}
KMVariables <- data.frame(time = fit$time,
                  n.risk = fit$n.risk,
                  n.event = fit$n.event,
                  n.censor = fit$n.censor,
                  surv = fit$surv,
                  upper = fit$upper,
                  lower = fit$lower
                  )
head(KMVariables)
```

### Make the graph of the function survfit()

```{r}
ggsurvplot(fit,
          pval = TRUE, conf.int = TRUE,
          risk.table = "abs_pct",  # absolute number and percentage at risk.
          risk.table.col = "strata", # Change risk table color by groups
          xlab = "Time in days",   # customize X axis label.
          break.time.by = 200,     # break X axis in time intervals by 200
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          legend.labs = c("Male", "Female"),    # change legend labels.
          ncensor.plot = TRUE,      # plot the number of censored subjects at time t
          palette = c("orange", "grey"))
```

Explanation : X-axis represents the time in days. Y-axis gives the probability of surviving. A vertical bar on the graph represents a censored event.

The median survival (50%) is 284 days for men and 426 days for women. Women seem surviving longer than men. This affirmation has to be confirmed with a Logrank Test. This test will allow to confirm or not that the two distributions have the same distribution curve.

With a p-value of 0.0014 the level of statistical significance is acceptable (below 0.05).

## Step 3 - Comparison of the two survival curves with Logrank test

-   Group 1 : Male
-   Group 2 : Female

Hypothesis :

-   H0 : Group 1 = Group 2
-   H1 : Group \<\> Group 2

The LOGRANK TEST is the method used to the two survival curves. The function to use in R is survdiff()

```{r}
surv_diff <- survdiff(Surv(time, status) ~ sex, data = lung)
surv_diff
```

Comments : a p-value of 0.001 reinforces confidence in the rejection of the null hypothesis and suggests that the differences observed between the two groups are very unlikely to be attributed to chance. The previous affirmation concerning female and male surviving is confirmed.

## Step 4 - Compute univariate Cox regression model

In univariate Cox regression only one factor is considered (in the example sex). The function coming from survival package used for Cox method is : coxph().

```{r}
univ.cox <- coxph(Surv(time, status) ~ sex, data = lung)
summary(univ.cox) # to show the report
```

Explanations of the indicators :

-   coef : this is the estimated coef Bêta form the Cox model.
-   exp(coef) : it is the Hazard Ratio (HR).
-   se(coef) : it is the std error of the coef Bêta.
-   z : z-score = coef/se(coef)
-   Pr(\>\|z\|) : the p-value associated with the z-score.
-   lower/upper : it is the confidence interval of the HR.
-   Concordance : is a measure of the predictive accuracy (1=perfect prediction/0.5=random).

Comments : An HR of 0.6193 suggests that, compared with the reference group, the female group associated with 0.62 has a reduced risk of events (e.g. death or disease progression).

A significantly low p-value, such as 0.00148, suggests a high probability that the association between sex and risk of lung cancer is not due to "chance".

Make the associated graph

```{r}
ggsurvplot(survfit(univ.cox,data=lung), 
           surv.median.line = "hv",
           break.time.by = 100,
           palette = "blue", 
           ggtheme = theme_minimal())
```

## Step 5 - Compute multivariate Cox regression model

We will compute the Cox method with all factors ; age, sex, ph.ecog, ph.karno, pat.karno, meal.cal, wt.loss and ph.ecog. We will see which factors are pertinent to keep.

```{r}
multi.cox1 <- coxph(Surv(time, status) ~ age + sex + ph.ecog + ph.karno + pat.karno + meal.cal + wt.loss, data =  lung)
summary(multi.cox1)
```

Remarks : Only the p-value of the covariates 'sex', 'ph.ecog' and 'ph.karno' are acceptable (below 0.05). The p-values of the other covariates are above 0.05 to be maintained in the following analysis.

We launch a new calculation with these three covariates.

```{r}
multi.cox2 <- coxph(Surv(time, status) ~ sex + ph.ecog + ph.karno, data =  lung)
summary(multi.cox2)
```

Comments : With a p-value above 0.05, the covariate "ph.karno" is no more relevant. A new computation with sex and ph.ecog is computed.

```{r}
multi.cox3 <- coxph(Surv(time, status) ~ sex + ph.ecog, data =  lung)
summary(multi.cox3)
```

Comments on the Hazard Ratio [exp(coef)] and p-value [Pr(\>\|z\|)]

With p-values lower than 0.05, the two covariates are relevant.

Concerning the covariate 'sex' the HR shows a different risk between male and female. Women have a lower death risk or lower progression of the cancer compared to men. This is confirmed with the survival curve in step 2.

Concerning the 'ph.ecog' covariate, the HR of 1.62 means an increase of the risk associated with a growth of the 'ph.ecog' indicator. An increase of 1 unit of the indicator 'ph.ecog', the risk associated increase of 62.01%

## Step 6 - Validate the number of covariates

6.1 Use of Likelihood Ratio Test (LRT)

It is a statistical test used to compare the goodness-of-fit of two nested models based on their likelihood functions.

We are going to compare the Cox Regression with 3 models :

-   sc1 = 'sex'only
-   sc2 = 'sex' + 'ph.ecog'
-   sc3 = 'sex' + 'ph.ecog' + 'ph.karno'

```{r}
lung <- na.omit(lung)  # Remove any rows with missing values
sc1 = coxph(Surv(time, status) ~ sex  , data =  lung)
sc2 = coxph(Surv(time, status) ~ sex + ph.ecog  , data =  lung)
sc3 = coxph(Surv(time, status) ~ sex + ph.ecog + ph.karno, data =  lung)
anova(sc1, sc2, sc3)
```

Comments: with a p-value of 0.000027, the second model, where we add ph.ecog improves the quality of the second model compare to the first one. However adding the 'age' does not offer a better model, as the p-value is 0.1016 (above 0.05).

6.2 Use the AIC model

```{r}
lung <- na.omit(lung)  # Remove any rows with missing values
sc1 = coxph(Surv(time, status) ~ sex  , data =  lung)
sc2 = coxph(Surv(time, status) ~ sex + ph.ecog  , data =  lung)
sc3 = coxph(Surv(time, status) ~ sex + ph.ecog + ph.karno, data =  lung)
fifi=list(sc1=sc1, sc2=sc2, sc3=sc3)
sapply(fifi, AIC)
```

Comments : We can see a decrease of the indicator between the model sc1 and model sc2. The improvement is not so strong between the model sc3 and the model sc2. Adding ph.karno does not bring any improvement.

## STEP 7 - PREDICTION

Predict and plot the survival curves at the age of 35, 55 and 85 for male only.

```{r}
cox_age=coxph(Surv(time, status) ~ age + sex  , data =  lung)
pred = survfit(cox_age, newdata = data.frame(age=c(35,55,85), sex=c(1,1,1)))

ggsurvplot(pred, data=lung,
          pval = FALSE, conf.int = FALSE,
          risk.table = FALSE,  # absolute number and percentage at risk.
          risk.table.col = "strata", # Change risk table color by groups
          xlab = "Time in days",   # customize X axis label.
          break.time.by = 100,     # break X axis in time intervals by 200
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          legend.labs = c("35y", "55y", "85y"),    # change legend labels.
          ncensor.plot = FALSE,      # plot the number of censored subjects at time t
          palette = c("orange", "grey","cyan"))
```

Comments : the older patient (85 years) has less surviving times (around 210 days) than a younger one (35 years - around 225 days).
